---
title: "Getting started with infernet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with infernet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(infernet)
library(manynet)
```

## Introduction

**infernet** provides inferential tests and regression models for network data.
It integrates with the **stocnet** ecosystem — accepting any
[manynet](https://stocnet.github.io/manynet/)-compatible network object and
any scalar network measure from **manynet**, **netrics**, or **migraph** as a
test statistic.

This vignette walks through a complete workflow:

1. Load a network from **manynet**
2. Run a conditional uniform graph (CUG) test with `test_random()`
3. Run a quadratic assignment procedure (QAP) test with `test_permutation()`
4. Run a network regression with `net_regression()`
5. Interpret the output

---

## 1. Load a network

We use the `ison_marvel_relationships` network bundled with **manynet**.
It is a signed network of relationships among Marvel comic characters.
We keep only the positive ("friend") ties and restrict to characters of
human origin.

```{r load-data}
marvel_friends <- manynet::to_unsigned(manynet::ison_marvel_relationships)
marvel_friends <- manynet::to_giant(marvel_friends) |>
  manynet::to_subgraph(PowerOrigin == "Human")

manynet::autographr(marvel_friends)
```

```{r network-summary}
manynet::net_nodes(marvel_friends)
manynet::net_ties(marvel_friends)
manynet::net_density(marvel_friends)
```

---

## 2. CUG test with `test_random()`

A **conditional uniform graph (CUG) test** asks: is the observed value of a
network statistic unusual compared to a distribution of values on random
networks of the same size?

Here we test whether the observed heterophily (tendency to tie to dissimilar
others) with respect to the `Attractive` attribute is higher or lower than
expected by chance.

`FUN` can be any function from **manynet**, **netrics**, or any package that
takes a network as its first argument and returns a scalar.

```{r cug-test}
set.seed(42)
cug_result <- test_random(
  marvel_friends,
  manynet::net_heterophily,
  attribute = "Attractive",
  times = 200
)
print(cug_result)
```

`Pr(X >= Obs)` is the fraction of random networks whose heterophily was at
least as large as the observed value.  A small value (e.g. < 0.05) suggests
the observed network is unusually homophilous.

```{r plot-cug}
plot(cug_result)
```

---

## 3. QAP test with `test_permutation()`

A **QAP test** uses permutations of the actual network rather than random
networks.  This is more conservative because it preserves the degree
distribution of the original graph.

```{r qap-test}
set.seed(42)
qap_result <- test_permutation(
  marvel_friends,
  manynet::net_heterophily,
  attribute = "Attractive",
  times = 200
)
print(qap_result)
plot(qap_result)
```

---

## 4. Network regression with `net_regression()`

`net_regression()` fits a multiple regression QAP (MRQAP) model to explain
tie weights or presence as a function of sender/receiver attributes and other
networks.

We use the `ison_networkers` dataset — a network of communication among
network researchers — and model tie strength (`weight`) as a function of
ego and alter citations and citation similarity.

```{r network-regression}
networkers <- manynet::ison_networkers |>
  manynet::to_subgraph(Discipline == "Sociology")

model1 <- net_regression(
  weight ~ ego(Citations) + alter(Citations) + sim(Citations),
  networkers,
  times = 200
)
print(model1)
```

### Reading the output

- **Estimate**: the OLS coefficient from the observed data.
- **Pr(<=t) / Pr(>=t) / Pr(>=|t|)**: permutation-based p-values.  The
  `qapspp` method (Dekker's double semi-partialling plus) is the default;
  these are reported in the `t`-statistic row.

A **positive `sim(Citations)` coefficient** means that researchers with
similar citation counts are more strongly connected.

### Using `net_regression_control()`

For advanced options, use `net_regression_control()` instead of writing a
raw list — all parameters are tab-completable:

```{r control-example, eval = FALSE}
ctrl <- net_regression_control(
  method  = "qapy",   # permute y only (faster, less conservative)
  family  = "gaussian",
  seed    = 123
)

model2 <- net_regression(
  weight ~ ego(Citations) + alter(Citations) + sim(Citations),
  networkers,
  times   = 200,
  control = ctrl
)
```

### Parallel execution

For large networks or many permutations, use `strategy = "multisession"`:

```{r parallel-example, eval = FALSE}
ctrl_par <- net_regression_control(strategy = "multisession")
model3   <- net_regression(
  weight ~ ego(Citations) + alter(Citations) + sim(Citations),
  networkers,
  times   = 1000,
  control = ctrl_par
)
```

---

## 5. Suppressing specification advice

When `sim()` or `same()` terms are present, **infernet** prints a reminder
to also include `ego()` / `alter()` terms for interpretability.  You can
silence this advice permanently with:

```{r advice-option, eval = FALSE}
options(infernet_advice = FALSE)
```

---

## Further reading

- See `?net_regression` and `?net_regression_control` for the full parameter
  reference.
- See `?test_random`, `?test_configuration`, and `?test_permutation` for the
  test functions.
- See `?net_from_edgelist` for converting long-format data to the matrix
  format expected by `net_regression()`.
- Visit the [stocnet GitHub organisation](https://github.com/stocnet/) for
  related packages: **manynet**, **netrics**, **migraph**.

## References

Krackhardt, David. 1988.
"Predicting with Networks: Nonparametric Multiple Regression Analysis of
Dyadic Data."
*Social Networks* 10(4): 359–381.
<https://doi.org/10.1016/0378-8733(88)90004-4>

Dekker, David, David Krackhardt, and Tom A. B. Snijders. 2007.
"Sensitivity of MRQAP tests to collinearity and autocorrelation conditions."
*Psychometrika* 72(4): 563–581.
<https://doi.org/10.1007/s11336-007-9016-1>
